Assignment:
Your assignment is to use your notes from class - along with help from classmates, UTAs, and me - to turn this script into a fleshed-out description of what is going on.

This is a substantial project - we'll work on it in steps over the rest of the unit. 

We are currently focused on the overall process and will cover the details over the rest of this unit.  Your answers can be brief.

Delete all the prompts like TODO() as you compete them.  Use RStudio's search function to see if you've missed any - there are a LOT!

When you've completed everything knit and upload it to RPubs.  A Canvas Assignment will be created where you can
1) post the link to your RPubs site
2) post the code

The next step after that will be announced later.




Add YAML header!!!
Give it a title


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# A complete bioinformatics workflow in R

<!-- TODO: create function to simplify the prepwork -->

**By**: Nathan L. Brouwer

# "Worked example: Building a phylogeny in R" 

## Introduction

Describe how phylogeneies can be used in biology
(readings will be assigned)




### Vocab 

Make a list of at least 10 vocab terms that are important (don't have to define)


### Key functions

Make a list of at least 5 key functions
Put in the format of package::function

## Software Preliminaires

Add the necessary calls to library() to load call packages
Indicate which packages cam from Bioconducotr, CRAN, and GitHub


### Load packages into memory

```{r, message= F, warning=F}
# github packages



# CRAN packages



# Bioconductor packages


```




## XXXXXXXXing macromolecular sequences

TODO: Fill in the XXXXXs and write a 1-2 sentence of what is going on here.  
Add the package that is where entrez_fetch is from using :: notation


```{r}
# Human shroom 3 (H. sapiens)
hShroom3 <- entrez_fetch(db = "protein", 
                          id = "NP_065910", 
                          rettype = "fasta")
```


TODO:explain what cat() is doing

```{r}
cat(hShroom3)
```


TODO: explain what this code chunk is doing

```{r}
# Mouse shroom 3a (M. musculus)
mShroom3a <- entrez_fetch(db = "protein", 
                          id = "AAF13269", 
                          rettype = "fasta")

# Human shroom 2 (H. sapiens)
hShroom2 <- entrez_fetch(db = "protein", 
                          id = "CAA58534", 
                          rettype = "fasta")


# Sea-urchin shroom
sShroom <- entrez_fetch(db = "protein", 
                          id = "XP_783573", 
                          rettype = "fasta")
```



TODO: Explain what this code chunk is doing 
```{r}
nchar(hShroom3)
nchar(mShroom3a)
nchar(sShroom)
nchar(hShroom2)
```



## Prepping macromolecular sequences

TODO: Explain what this function does

```{r}
fasta_cleaner
```


TODO: explain how to add the function to your R session even if you can't download compbio4all

```{r}
fasta_cleaner <- function(fasta_object, parse = TRUE){

  fasta_object <- sub("^(>)(.*?)(\\n)(.*)(\\n\\n)","\\4",fasta_object)
  fasta_object <- gsub("\n", "", fasta_object)

  if(parse == TRUE){
    fasta_object <- stringr::str_split(fasta_object,
                                       pattern = "",
                                       simplify = FALSE)
  }

  return(fasta_object[[1]])
}

```



TODO: briefly explain what this code chunk is doing

```{r}
hShroom3  <- fasta_cleaner(hShroom3,  parse = F)
mShroom3a <- fasta_cleaner(mShroom3a, parse = F)
hShroom2  <- fasta_cleaner(hShroom2,  parse = F)
sShroom   <- fasta_cleaner(sShroom,   parse = F)
```




```{r}
hShroom3
```


## XXXXXXXXing sequences

TODO: give this a title.  Explain what code below is doing

```{r}
# add necessary function
align.h3.vs.m3a <- Biostrings::           (
                  hShroom3,
                  mShroom3a)
```


TODO: In 1-2 sentence explain what this object shows
```{r}
align.h3.vs.m3a
```




TODO: explain what this is showing
```{r}
# add necessary function
Biostrings::    (align.h3.vs.m3a)
```




TODO: briefly explain what is going on here versus the previous code chunk
```{r}
align.h3.vs.h2 <- Biostrings::pairwiseAlignment(
                  hShroom3,
                  hShroom2)
```



TODO: explain what is going on here and compare and contrast with previous ouput
```{r}
score(align.h3.vs.h2)
```



TODO: briefly explian the difference between the output of score() and pid() (can be very brief - we'll get into the details later)
```{r}
Biostrings::pid(align.h3.vs.h2)
```





## The shroom family of genes

TODO: briefly explain why I have this whole table here



```{r}
shroom_table <- c("CAA78718" , "X. laevis Apx" ,         "xShroom1",
            "NP_597713" , "H. sapiens APXL2" ,     "hShroom1",
            "CAA58534" , "H. sapiens APXL",        "hShroom2",
            "ABD19518" , "M. musculus Apxl" ,      "mShroom2",
            "AAF13269" , "M. musculus ShroomL" ,   "mShroom3a",
            "AAF13270" , "M. musculus ShroomS" ,   "mShroom3b",
            "NP_065910", "H. sapiens Shroom" ,     "hShroom3",
            "ABD59319" , "X. laevis Shroom-like",  "xShroom3",
            "NP_065768", "H. sapiens KIAA1202" ,   "hShroom4a",
            "AAK95579" , "H. sapiens SHAP-A" ,     "hShroom4b",
            #"DQ435686" , "M. musculus KIAA1202" ,  "mShroom4",
            "ABA81834" , "D. melanogaster Shroom", "dmShroom",
            "EAA12598" , "A. gambiae Shroom",      "agShroom",
            "XP_392427" , "A. mellifera Shroom" ,  "amShroom",
            "XP_783573" , "S. purpuratus Shroom" , "spShroom") #sea urchin
```



TODO: write a short sentence explaining what this next code chunk will do, then annotate each line with what was done.

```{r}
# convert to XXXXXXXXXC
shroom_table_matrix <- matrix(shroom_table,
                                  byrow = T,
                                  nrow = 14)
# convert to XXXXXXXXXC
shroom_table <- data.frame(shroom_table_matrix, 
                     stringsAsFactors = F)

# XXXXXXXXXC columns
names(shroom_table) <- c("accession", "name.orig","name.new")

# Create simplified species names
shroom_table$spp <- "Homo"
shroom_table$spp[grep("laevis",shroom_table$name.orig)] <- "Xenopus"
shroom_table$spp[grep("musculus",shroom_table$name.orig)] <- "Mus"
shroom_table$spp[grep("melanogaster",shroom_table$name.orig)] <- "Drosophila"
shroom_table$spp[grep("gambiae",shroom_table$name.orig)] <- "mosquito"
shroom_table$spp[grep("mellifera",shroom_table$name.orig)] <- "bee"
shroom_table$spp[grep("purpuratus",shroom_table$name.orig)] <- "sea urchin"

```


TODO: in a brief sentence explain what this is doing

```{r}
shroom_table
```

## XXXXXing multiple sequences 


TODO: in a brief sentence explain what the $ allows us to do

```{r}
shroom_table$accession
```



TODO: briefly explain what this chunk is doing and add the correct function
```{r}
# add necessary function
shrooms <-          (db = "protein", 
                          id = shroom_table$accession, 
                          rettype = "fasta")
```



TODO: in a very brief sentence explain what this is doing.
```{r, eval = F}
cat(shrooms)
```




TODO: in a  brief sentence explain what this is doing and if/how its different from the previous code chunks
```{r}
shrooms_list <- entrez_fetch_list(db = "protein", 
                          id = shroom_table$accession, 
                          rettype = "fasta")
```



TODO: briefly explain what I am doing this
```{r}
length(shrooms_list)
```



TODO: briefly explain what I am doing this.  We will get into the details of for() loops in R later in the semester.

```{r}
for(i in 1:length(shrooms_list)){
  shrooms_list[[i]] <- fasta_cleaner(shrooms_list[[i]], parse = F)
}
```


TODO: summarize what is going on in this code chunk, then annotate each line of code with what its doing

```{r}
# XXXXXXXXCX
shrooms_vector <- rep(NA, length(shrooms_list))

# XXXXXXXXCX
for(i in 1:length(shrooms_vector)){
  shrooms_vector[i] <- shrooms_list[[i]]
}

#  XXXXXXXXCX
names(shrooms_vector) <- names(shrooms_list)
```



TODO: explain what this is doing then add the necessary function.
```{r}
# add necessary function
shrooms_vector_ss <- Biostrings::        (shrooms_vector)
```


## MSA


TODO: briefly summarize what this section of the document will do.  
Readings will be assigned to explain what MSAs are.




### Building an XXXXXXXX (MSA)



TODO: briefly explain what this chunk does, then add the necessary function.

```{r}
# add necessary function
shrooms_align <-      (shrooms_vector_ss,
                     method = "ClustalW")
```


### Viewing an MSA

TODO: briefly summarize what this section will do.


#### Viewing an MSA in R


TODO: Briefly summarize what output is shown below 

```{r}
shrooms_align
```



TODO: briefly explain what is being done in this chunk.  This is tricky (and annoying) so do your best



```{r}
# WHAT IS THE LINE BELOW DOING? (its tricky - do your best)
class(shrooms_align) <- "AAMultipleAlignment"

# WHAT IS THE LINE BELOW DOING? This is simpler
shrooms_align_seqinr <- msaConvert(shrooms_align, type = "seqinr::alignment")
```




TODO: what is the output this produces
```{r, eval = F}
print_msa(alignment = shrooms_align_seqinr, 
          chunksize = 60)
```

#### Displaying an MSA XXXXXXXX



TODO: explain this output and how its differnet from the prevoius

```{r}
## add necessary function
ggmsa::     (shrooms_align,   # shrooms_align, NOT shrooms_align_seqinr
      start = 2000, 
      end = 2100) 
```



#### Saving an MSA as PDF


TODO: explain what this command is doing.
Add the package the function is coming from using :: notation
This may not work for everyone.  If its not working you can comment it out.


```{r, eval = F}
msaPrettyPrint(shrooms_align,             # alignment
               file = "shroom_msa.pdf",   # file name
               y=c(2000, 2100),           # range
               askForOverwrite=FALSE)
```


TODO: explain what this command is doing

```{r}
getwd()
```




## Genetic XXXXXXXXX

TODO: give this section a title and briefly explain what is going on below.

```{r}
### add necessary function
shrooms_dist <- seqinr::       (shrooms_align_seqinr, 
                                       matrix = "identity")


```



TODO:  briefly explain what is going on below.

```{r}
### add necessary function
shrooms_dist_rounded <-       (shrooms_dist,
                              digits = 3)
```


TODO:  briefly explain what is being shown.

```{r}
shrooms_dist_rounded
```

## Phylognetic XXXXXX (finally!)



TODO: explain what this section is doing

```{r}
### add necessary function
tree <-        (shrooms_dist)
```

### Plotting XXXXXX

TODO: explain what this section is doing


```{r}
# plot tree
plot.phylo(tree, main="Phylogenetic Tree", 
            type = "unrooted", 
            use.edge.length = F)

# add label
mtext(text = "Shroom family gene tree - unrooted, no branch lengths")
```



TODO: briefly explain what this section is doing
How is it different from the previous code chunk?
(do your best - we'll cover the details later)

```{r}
# plot tree
plot.phylo(tree, main="Phylogenetic Tree", 
            use.edge.length = F)

# add label
mtext(text = "Shroom family gene tree - rooted, no branch lenths")
```



TODO: briefly explain what this section is doing
How is it different from the previous code chunk?
(do your best - we'll cover the details later)


```{r}
# plot tree
plot.phylo(tree, main="Phylogenetic Tree", 
            use.edge.length = T)

# add label
mtext(text = "Shroom family gene tree - rooted, with branch lenths")
```


TODO: briefly explain what this section is doing
How is it different from the previous code chunk?
Focus on what graphical elements (also this code sometimes doesn't make a great plot - I need to update it) 



```{r fig.width=6}
plot(tree, main="Phylogenetic Tree")
mtext(text = "Shroom family gene tree")

x <- 0.551
x2 <- 0.6

# label Shrm 3
segments(x0 = x, y0 = 1, 
         x1 = x, y1 = 4,
         lwd=2)
text(x = x*1.01, y = 2.5, "Shrm 3",adj = 0)

segments(x0 = x, y0 = 5, 
         x1 = x, y1 = 6,
         lwd=2)
text(x = x*1.01, y = 5.5, "Shrm 2",adj = 0)

segments(x0 = x, y0 = 7, 
         x1 = x, y1 = 9,
         lwd=2)
text(x = x*1.01, y = 8, "Shrm 1",adj = 0)

segments(x0 = x, y0 = 10, 
         x1 = x, y1 = 13,
         lwd=2)
text(x = x*1.01, y = 12, "Shrm ?",adj = 0)


segments(x0 = x, y0 = 14, 
         x1 = x, y1 = 15,
         lwd=2)
text(x = x*1.01, y = 14.5, "Shrm 4",adj = 0)


segments(x0 = x2, y0 = 1, 
         x1 = x2, y1 = 6,
         lwd=2)

segments(x0 = x2, y0 = 7, 
         x1 = x2, y1 = 9,
         lwd=2)

segments(x0 = x2, y0 = 10, 
         x1 = x2, y1 = 15,
         lwd=2)

```




