---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Hidden Markov Models

```{r  eval = T}
library(compbio4all) 
```

**By**: Avril Coghlan.  

**Adapted, edited and expanded**: Nathan Brouwer under the Creative Commons 3.0 Attribution License [(CC BY 3.0)](https://creativecommons.org/licenses/by/3.0/) with assistance from Havannah Tung.


**NOTE: this chapter has not yet been significaly revised**.

## A multinomial model of DNA sequence evolution

The simplest model of DNA sequence evolution assumes that the sequence has been produced by a random process that randomly chose any of the four nucleotides at each position in the sequence, where the probability of choosing any one of the four nucleotides depends on a predetermined probability distribution. That is, the four nucleotides are chosen with pA, pC, pG, and pT respectively. This is known as the multinomial sequence model.

A multinomial model for DNA sequence evolution has four parameters: the probabilities of the four nucleotides pA, pC, pG, and pT. For example, say we may create a multinomial model where pA=0.2, pC=0.3, pG=0.3, and pT=0.2. This means that the probability of choosing a A at any particular sequence position is set to be 0.2, the probability of choosing a C is 0.3, of choosing a G is 0.3, and of choosing a T is 0.2. Note that pA + pC + pG + pT = 1, as the sum of the probabilities of the four different types of nucleotides must be equal to 1, as there are only four possible types of nucleotide.

The multinomial sequence model is like having a roulette wheel that is divided into four different slices labeled “A”, “T”, “G” and “C”, where the pA, pT, pG and pC are the fractions of the wheel taken up by the slices with these four labels. If you spin the arrow attached to the center of the roulette wheel, the probability that it will stop in the slice with a particular label (eg. the slice labeled “A”) only depends on the fraction of the wheel taken up by that slice (pA here; SEE THE PICTURE BELOW).


## Generating a DNA sequence using a multinomial model

We can use R to generate a DNA sequence using a particular multinomial model. First we need to set the values of the four parameters of the multinomial model, the probabilities pA, pC, pG, and pT of choosing the nucleotides A, C, G and T, respectively, at a particular position in the DNA sequence. For example, say we decide to set pA=0.2, pC=0.3, pG=0.3, and pT=0.2. We can use the function sample() in R to generate a DNA sequence of a certain length, by selecting a nucleotide at each position according to this probability distribution:

Define the alphabet of nucleotides
```{r}
nucleotides    <- c("A", "C", "G", "T") 
```

Set the values of the probabilities
```{r}
probabilities1 <- c(0.2, 0.3, 0.3, 0.2) 
```

Set the length of the sequence
```{r}
seqlength      <- 30    
```

Generate a sequence
```{r}
sample(nucleotides, 
       seqlength,
       rep=TRUE,
       prob=probabilities1) 
```

If you look at the help page for the function, you will find that its inputs are the vector to sample from (nucleotides here), the size of the sample (seqlength here), and a vector of probabilities for obtaining the elements of the vector being sampled (probabilities1 here). If we use the sample() function to generate a sequence again, it will create a different sequence using the same multinomial model:

```{r}
# Generate another sequence
sample(nucleotides, 
       seqlength, 
       rep=TRUE, 
       prob=probabilities1) 
```

In the same way, we can generate a sequence using a different multinomial model, where pA=0.1, pC=0.41, pG=0.39, and pT=0.1:
```{r}
# Set the values of the probabilities for the new model
probabilities2 <- c(0.1, 0.41, 0.39, 0.1) 

# Generate a sequence
sample(nucleotides, seqlength, rep=TRUE, prob=probabilities2) 
```

As you would expect, the sequences generated using this second multinomial model have a higher fraction of Cs and Gs compared to the sequences generated using the first multinomial model above. This is because pC and GT are higher for this second model than for the first model (pC=0.41 and GT=0.39 in the second model, versus pC=0.3 and GT=0.3 in the first model). That is, in the second multinomial model we are using a roulette wheel that has large slices labeled “C” and “G”, while in the first multinomial model we were using a roulette wheel with relatively smaller slices labeled “C” and “G” (SEE THE PICTURE BELOW).


## A Markov model of DNA sequence evolution

A multinomial model of DNA sequence evolution is a good model of the evolution of many DNA sequences. However, for some DNA sequences, a multinomial model is not an accurate representation of how the sequences have evolved. One reason is that a multinomial model assumes that each part of the sequence (eg. the first 100 nucleotides of the sequence, the second 100 nucleotides, the third 100 nucleotides, etc.) have the same frequency of each type of nucleotide (the same pA, pC, pG, and pT), and this may not be true for a particular DNA sequence if there are considerable differences in nucleotide frequencies in different parts of the sequence.

Another assumption of a multinomial model of DNA sequence evolution is that the probability of choosing a particular nucleotide (eg. “A”) at a particular position in the sequence only depends on the predetermined frequency of that nucleotide (pA here), and does not depend at all on the nucleotides found at adjacent positions in the sequence. This assumption holds true for many DNA sequences. However, for some DNA sequences, it is not true, because the probability of finding a particular nucleotide at a particular position in the sequence does depend on what nucleotides are found at adjacent positions in the sequence. In this case, a different type of DNA sequence model called a Markov sequence model is a more accurate representation of the evolution of the sequence.

A Markov sequence model assumes that the sequence has been produced by a process that chose any of the four nucleotides in the sequence, where the probability of choosing any one of the four nucleotides at a particular position depends on the nucleotide chosen for the previous position. That is, if “A” was chosen at the previous position, then the probability of choosing any one of the four nucleotides at the current position depends on a predetermined probability distribution. That is, given that “A” was chosen at the previous position, the four nucleotides are chosen at the current position with probabilities of pA, pC, pG, and pT of choosing “A”, “C”, “G”, or “T”, respectively (eg. pA=0.2, pC=0.3, pG=0.3, and pT=0.2). In contrast, if “C” was chosen at the previous position, then the probability of choosing any one of the four nucleotides at the current position depends on a different predetermined probability distribution, that is, the probabilities of choosing “A”, “C”, “G”, or “T” at the current position are now different (eg. pA=0.1, pC=0.41, pG=0.39, and pT=0.1).

A Markov sequence model is like having four different roulette wheels, labeled “afterA”, “afterT”, “afterG”, and “afterC”, for the cases when “A”, “T”, “G”, or “C” were chosen at the previous position in a sequence, respectively. Each of the four roulette wheels has four slices labeled “A”, “T”, “G”, and “C”, but in each roulette wheel a different fraction of the wheel is taken up by the four slices. That is, each roulette wheel has a different pA, pT, pG and pC. If we are generating a new DNA sequence using a Markov sequence model, to decide what nucleotide to choose at a particular position in the sequence, you spin the arrow at the center of a roulette wheel, and see in which slice the arrow stops. There are four roulette wheels, and the particular roulette wheel we use at a particular position in the sequence depends on the nucleotide chosen for the previous position in the sequence. For example, if “T” was chosen at the previous position, we use the “afterT” roulette wheel to choose the nucleotide for the current position. The probability of choosing a particular nucleotide at the current position (eg. “A”) then depends on the fraction of the “afterT” roulette wheel taken up by the the slice labeled with that nucleotide (pA here; SEE THE PICTURE BELOW).


## The transition matrix for a Markov model

A multinomial model of DNA sequence evolution just has four parameters: the probabilities pA, pC, pG, and pT. In contrast, a Markov model has many more parameters: four sets of probabilities pA, pC, pG, and pT, that differ according to whether the previous nucleotide was “A”, “G”, “T” or “C”. The symbols pAA, pAC, pAG, and pAT are usually used to represent the four probabilities for the case where the previous nucleotide was “A”, the symbols pCA, pCC, pCG, and pCT for the case when the previous nucleotide was “C”, and so on.

It is common to store the probability parameters for a Markov model of a DNA sequence in a square matrix, which is known as a Markov transition matrix. The rows of the transition matrix represent the nucleotide found at the previous position in the sequence, while the columns represent the nucleotides that could be found at the current position in the sequence. In R, you can create a matrix using the *matrix()* command, and the rownames() and colnames() functions can be used to label the rows and columns of the matrix. For example, to create a transition matrix, we type:

Define the alphabet of nucleotides
```{r}
nucleotides         <- c("A", "C", "G", "T") 
```


Set the values of the probabilities, where the previous nucleotide was "A"
```{r}
afterAprobs <- c(0.2, 0.3, 0.3, 0.2)  
```

Set the values of the probabilities, where the previous nucleotide was "C"
```{r}
afterCprobs <- c(0.1, 0.41, 0.39, 0.1)       
```

Set the values of the probabilities, where the previous nucleotide was "G"
```{r}
afterGprobs <- c(0.25, 0.25, 0.25, 0.25)   
```

Set the values of the probabilities, where the previous nucleotide was "T"
```{r}
afterTprobs <- c(0.5, 0.17, 0.17, 0.17)
```

Create a 4 x 4 matrix
```{r}
mytransitionmatrix <- matrix(c(afterAprobs, 
                               afterCprobs, 
                               afterGprobs, 
                               afterTprobs),
                             4, 4, byrow = TRUE)

rownames(mytransitionmatrix) <- nucleotides
colnames(mytransitionmatrix) <- nucleotides
```


Print out the transition matrix
```{r}
mytransitionmatrix    
```

Rows 1, 2, 3 and 4 of the transition matrix give the probabilities pA, pC, pG, and pT for the cases where the previous nucleotide was “A”, “C”, “G”, or “T”, respectively. That is, the element in a particular row and column of the transition matrix (eg. the row for “A”, column for “C”) holds the probability (pAC) of choosing a particular nucleotide (“C”) at the current position in the sequence, given that was a particular nucleotide (“A”) at the previous position in the sequence.


## Generating a DNA sequence using a Markov model

Just as you can generate a DNA sequence using a particular multinomial model, you can generate a DNA sequence using a particular Markov model. When you are generating a DNA sequence using a Markov model, the nucleotide chosen at each position at the sequence depends on the nucleotide chosen at the previous position. As there is no previous nucleotide at the first position in the new sequence, we need to define the probabilities of choosing “A”, “C”, “G” or “T” for the first position. The symbols piA, piC, piG, and piT are used to represent the probabilities of choosing “A”, “C”, “G”, or “T” at the first position.

We will use an *R* function `generate_markov_seq()` (Coghlan 2011) to generate a DNA sequence using a particular Markov model.

The function `generate_markov_seq()` takes as its arguments (inputs) the transition matrix for the particular Markov model; a vector containing the values of piA, piC, piG, and piT; and the length of the DNA sequence to be generated.

The probabilities of choosing each of the four nucleotides at the first position in the sequence are piA, piC, piG, and piT. The probabilities of choosing each of the four nucleotides at the second position in the sequence depend on the particular nucleotide that was chosen at the first position in the sequence. The probabilities of choosing each of the four nucleotides at the third position depend on the nucleotide chosen at the second position, and so on.

We can use the `generate_markov_seq()` function to generate a sequence using a particular Markov model. For example, to create a sequence of 30 nucleotides using the Markov model described in the transition matrix mytransitionmatrix, using uniform starting probabilities (ie. piA = 0.25, piC = 0.25, piG = 0.25, and piT = 0.25) , we type:
```{r}
myinitialprobs <- c(0.25, 0.25, 0.25, 0.25)

generate_markov_seq(mytransitionmatrix, myinitialprobs, 30)
```

As you can see, there are many “A”s after “T”s in the sequence. This is because pTA has a high value (0.5) in the Markov transition matrix mytransitionmatrix. Similarly, there are few “A”s or “T”s after “C”s, which is because pCA and pCT have low values (0.1) in this transition matrix.


## A Hidden Markov Model of DNA sequence evolution

In a Markov model, the nucleotide at a particular position in a sequence depends on the nucleotide found at the previous position. In contrast, in a Hidden Markov model (HMM), the nucleotide found at a particular position in a sequence depends on the state at the previous nucleotide position in the sequence. The state at a sequence position is a property of that position of the sequence, for example, a particular HMM may model the positions along a sequence as belonging to either one of two states, “GC-rich” or “AT-rich”. A more complex HMM may model the positions along a sequence as belonging to many different possible states, such as “**promoter**”, “**exon**”, “**intron**”, and “**intergenic DNA**”.

A HMM is like having several different roulette wheels, one roulette wheel for each state in the HMM, for example, a “GC-rich” and an “AT-rich” roulette wheel. Each of the roulette wheels has four slices labeled “A”, “T”, “G”, and “C”, and in each roulette wheel a different fraction of the wheel is taken up by the four slices. That is, the “GC-rich” and “AT-rich” roulette wheels have different pA, pT, pG and pC values. If we are generating a new DNA sequence using a HMM, to decide what nucleotide to choose at a particular sequence position, we spin the arrow of a particular roulette wheel, and see in which slice it stops.

How do we decide which roulette wheel to use? Well, if there are two roulette wheels, we tend to use the same roulette wheel that we used to choose the previous nucleotide in the sequence, but there is also a certain small probability of switching to the other roulette wheel. For example, if we used the “GC-rich” roulette wheel to choose the previous nucleotide in the sequence, there may be a 90% chance that we will use the “GC-rich” roulette wheel again to choose the nucleotide at the current position, but a 10% chance that we will switch to using the “AT-rich” roulette wheel to choose the nucleotide at the current position. Likewise, if we used the “AT-rich” roulette wheel to choose the nucleotide at the previous position, there may be 70% chance that we will use the “AT-rich” wheel again at this position, but a 30% chance that we will switch to using the “GC-rich” roulette wheel to choose the nucleotide at this position.


## The transition matrix and emission matrix for a HMM

A HMM has two important matrices that hold its parameters. The first is the HMM transition matrix, which contains the probabilities of switching from one state to another. For example, in a HMM with two states, an AT-rich state and a GC-rich state, the transition matrix will hold the probabilities of switching from the AT-rich state to the GC-rich state, and of switching from the GC-rich state to the AT-rich state. For example, if the previous nucleotide was in the AT-rich state there may be a probability of 0.3 that the current nucleotide will be in the GC-rich state, and if the previous nucleotide was in the GC-rich state there may be a probability of 0.1 that the current nucleotide will be in the AT-rich state:

Define the names of the states
```{r}
states              <- c("AT-rich", "GC-rich")
```

Set the probabilities of switching states, where the previous state was "AT-rich"
```{r}
ATrichprobs         <- c(0.7, 0.3)  
```

Set the probabilities of switching states, where the previous state was "GC-rich"
```{r}
GCrichprobs         <- c(0.1, 0.9)   
```

Create a 2 x 2 matrix
```{r}
thetransitionmatrix <- matrix(c(ATrichprobs, 
                                GCrichprobs), 
                              2, 2, 
                              byrow = TRUE)

rownames(thetransitionmatrix) <- states
colnames(thetransitionmatrix) <- states
```


Print out the transition matrix
```{r}
thetransitionmatrix   
```

There is a row in the transition matrix for each of the possible states at the previous position in the nucleotide sequence. For example, in this transition matrix, the first row corresponds to the case where the previous position was in the “AT-rich” state, and the second row corresponds to the case where the previous position was in the “GC-rich” state. The columns give the probabilities of switching to different states at the current position. For example, the value in the second row and first column of the transition matrix above is 0.1, which is the probability of switching to the AT-rich state, if the previous position of the sequence was in the GC-rich state.

The second important matrix is the HMM emission matrix, which holds the probabilities of choosing the four nucleotides “A”, “C”, “G”, and “T”, in each of the states. In a HMM with an AT-rich state and a GC-rich state, the emission matrix will hold the probabilities of choosing each of the four nucleotides “A”, “C”, “G” and “T” in the AT-rich state (for example, pA=0.39, pC=0.1, pG=0.1, and pT=0.41 for the AT-rich state), and the probabilities of choosing “A”, “C”, “G”, and “T” in the GC-rich state (for example, pA=0.1, pC=0.41, pG=0.39, and pT=0.1 for the GC-rich state).

```{r}
 # Define the alphabet of nucleotides
nucleotides         <- c("A", "C", "G", "T")  

# Set the values of the probabilities, for the AT-rich state
ATrichstateprobs    <- c(0.39, 0.1, 0.1, 0.41)

# Set the values of the probabilities, for the GC-rich state
GCrichstateprobs    <- c(0.1, 0.41, 0.39, 0.1) 

# Create a 2 x 4 matrix
theemissionmatrix <- matrix(c(ATrichstateprobs, 
                              GCrichstateprobs), 
                            2, 4, 
                            byrow = TRUE) 

rownames(theemissionmatrix) <- states
colnames(theemissionmatrix) <- nucleotides

# Print out the emission matrix
theemissionmatrix                              
```

There is a row in the emission matrix for each possible state, and the columns give the probabilities of choosing each of the four possible nucleotides when in a particular state. For example, the value in the second row and third column of the emission matrix above is 0.39, which is the probability of choosing a “G” when in the “GC-rich state” (ie. when using the “GC-rich” roulette wheel).


## Generating a DNA sequence using a HMM

The following function `generate_hmm_seq()` can be used to generate a DNA sequence using a particular HMM. As its arguments (inputs), it requires the parameters of the HMM: the HMM transmission matrix and HMM emission matrix.

When you are generating a DNA sequence using a HMM, the nucleotide is chosen at each position depending on the state at the previous position in the sequence. As there is no previous nucleotide at the first position in the sequence, the function `generate_hmm_seq()` also requires the probabilities of the choosing each of the states at the first position (eg. piAT-rich and piGC-rich being the probability of the choosing the “AT-rich” or “GC-rich” states at the first position for a HMM with these two states).

We can use the `generate_hmm_seq()` function to generate a sequence using a particular HMM. For example, to create a sequence of 30 nucleotides using the HMM with “AT-rich” and “GC-rich” states described in the transition matrix thetransitionmatrix, the emission matrix theemissionmatrix, and uniform starting probabilities (ie. piAT-rich = 0.5, piGC-rich = 0.5), we type:

```{r}
theinitialprobs <- c(0.5, 0.5)

generate_hmm_seq(thetransitionmatrix, 
                 theemissionmatrix, 
                 theinitialprobs, 30)
```

As you can see, the nucleotides generated by the GC-rich state are mostly but not all “G”s and “C”s (because of the high values of pG and pC for the GC-rich state in the HMM emission matrix), while the nucleotides generated by the AT-rich state are mostly but not all “A”s and “T”s (because of the high values of pT and pA for the AT-rics state in the HMM emission matrix).

Furthermore, there tends to be runs of nucleotides that are either all in the GC-rich state or all in the AT-rich state, as the transition matrix specifies that the probabilities of switching from the AT-rich to GC-rich state (probability 0.3), or GC-rich to AT-rich state (probability 0.1) are relatively low.


## Inferring the states of a HMM that generated a DNA sequence

If we have a HMM with two states, “GC-rich” and “AT-rich”, and we know the transmission and emission matrices of the HMM, can we take some new DNA sequence, and figure out which state (GC-rich or AT-rich) is the most likely to have generated each nucleotide position in that DNA sequence? This is a common problem in bioinformatics. It is called the problem of finding the most probable state path, as it essentially consists of assigning the most likely state to each position in the DNA sequence. The problem of finding the most probable state path is also sometimes called segmentation. For example, give a DNA sequence of 1000 nucleotides, you may wish to use your HMM to segment the sequence into blocks that were probably generated by the “GC-rich” state or by the “AT-rich” state.

The problem of finding the most probable state path given a HMM and a sequence (ie. the problem of segmenting a sequence using a HMM), can be solved by an algorithm called the Viterbi algorithm. As its output, the Viterbi algorithm gives for each nucleotide position in a DNA sequence, the state of your HMM that most probably generated the nucleotide in that position. For example, if you segmented a particular DNA sequence of 1000 nucleotides using a HMM with “AT-rich” and “GC-rich” states, the Viterbi algorithm may tell you that nucleotides 1-343 were most probably generated by the AT-rich state, nucleotides 344-900 were most probably generated by the GC-rich state, and 901-1000 were most probably generated by the AT-rich state.

The function `viterbi()` (Coghlan 2011) is a function for the Viterbi algorithm.  The `viterbi()` function requires a second function `make_viterbi_mat()`.  Given a HMM, and a particular DNA sequence, you can use the `viterbi()` function to find the state of that HMM that was most likely to have generated the nucleotide at each position in the DNA sequence:
```{r}
myseq <- c("A", "A", "G", "C", "G", "T", "G", "G", "G", "G", "C", "C", "C", "C", "G", "G", "C", "G", "A", "C", "A", "T", "G", "G", "G", "G", "T", "G", "T", "C")

compbio4all::viterbi(myseq, thetransitionmatrix, theemissionmatrix)
```


## A Hidden Markov Model of protein sequence evolution

We have so far talked about using HMMs to model DNA sequence evolution. However, it is of course possible to use HMMs to model protein sequence evolution. When using a HMM to model DNA sequence evolution, we may have states such as “AT-rich” and “GC-rich”. Similarly, when using a HMM to model protein sequence evolution, we may have states such as “hydrophobic” and “hydrophilic”. In a protein HMM with “hydrophilic” and “hydrophilic” states, the “hydrophilic” HMM will have probabilities pA, pR, pC... of choosing each of the 20 amino acids alanine (A), arginine (R), cysteine (C), etc. when in that state. Similarly, the “hydrophilic” state will have different probabilities pA, pR, pC... of choosing each of the 20 amino acids. The probability of choosing a hydrophobic amino acid such as alanine will be higher in the “hydrophobic” state than in the “hydrophilic” state (ie. pA of the “hydrophobic” state will be higher than the pA of of the “hydrophilic” state, where A represents alanine here). A HMM of protein sequence evolution also defines a certain probability of switching from the “hydrophilic” state to the “hydrophobic” state, and a certain probability of switching from the “hydrophobic” state to the “hydrophilic” state.


## Summary

In this practical, you will have learned to use the following R functions:

1. `numeric()` for making a vector for storing numbers
1. `character()` for making a vector for storing characters
1. `matrix()` for making a matrix variable
1. `rownames()` for assigning names to the rows of a matrix variable
1. `colnames()` for assigning names to the columns of a matrix variable
1. `sample()` for making a random sample of numbers from a vector of numbers

All of these functions belong to the standard installation of *R*.


## Links and Further Reading

Some links are included here for further reading, which will be especially useful if you need to use the R package for your project or assignments.

For background reading on multinomial models, Markov models, and HMMs, it is recommended to read Chapters 1 and 4 of Introduction to Computational Genomics: a case studies approach by Cristianini and Hahn (Cambridge University Press; www.computational-genomics.net/book/).

There is also a very nice chapter on “Markov Models” in the book Applied statistics for bioinformatics using R by Krijnen (available online at cran.r-project.org/doc/contrib/Krijnen-IntroBioInfStatistics.pdf).


## Exercises

Answer the following questions, using the R package. For each question, please record your answer, and what you typed into R to get this answer.

1. In a previous practical, you saw that the Bacteriophage lambda genome sequence (NCBI accession NC_001416) has long stretches of either very GC-rich (mostly in the first half of the genome) or very AT-rich sequence (mostly in the second half of the genome). Use a HMM with two different states (“AT-rich” and “GC-rich”) to infer which state of the HMM is most likely to have generated each nucleotide position in the Bacteriophage lambda genome sequence. For the AT-rich state, set pA= 0.27, pC= 0.2084, pG= 0.198, and pT= 0.3236. For the GC-rich state, set pA= 0.2462, pC= 0.2476, pG= 0.2985, and pT= 0.2077. Set the probability of switching from the AT-rich state to the GC-rich state to be 0.0002, and the probability of switching from the GC-rich state to the AT-rich state to be 0.0002. What is the most probable state path?
1. Given a HMM with four different states (“A-rich”, “C-rich”, “G-rich” and “T-rich”), infer which state of the HMM is most likely to have generated each nucleotide position in the Bacteriophage lambda genome sequence. For the A-rich state, set pA= 0.3236, pC= 0.2084, pG= 0.198, and pT= 0.27. For the C-rich state, set pA= 0.2462, pC= 0.2985, pG= 0.2476, and pT= 0.2077. For the G-rich state, set pA= 0.2462, pC= 0.2476, pG= 0.2985, and pT= 0.2077. For the T-rich state, set pA= 0.27, pC= 0.2084, pG= 0.198, and pT= 0.3236. Set the probability of switching from the A-rich state to any of the three other states to be 6.666667e-05. Likewise, set the probability of switching from the C-rich/G-rich/T-rich state to any of the three other states to be 6.666667e-05. What is the most probable state path? Do you find differences between these results and the results from simply using a two-state HMM (as in Q1)?
1. Make a two-state HMM to model protein sequence evolution, with “hydrophilic” and “hydrophobic” states. For the hydrophilic state, set pA= 0.02, pR= 0.068, pN= 0.068, pD= 0.068, pC= 0.02, pQ= 0.068, pE= 0.068, pG= 0.068, pH= 0.068, pI= 0.012, pL= 0.012, pK= 0.068, pM= 0.02, pF= 0.02, pP= 0.068, pS= 0.068, pT= 0.068, pW= 0.068, pY= 0.068, and pV= 0.012. For the hydrophobic state, set pA= 0.114, pR= 0.007, pN= 0.007, pD= 0.007, pC= 0.114, pQ= 0.007, pE= 0.007, pG= 0.025, pH= 0.007, pI= 0.114, pL= 0.114, pK= 0.007, pM= 0.114, pF= 0.114, pP= 0.025, pS= 0.026, pT= 0.026, pW= 0.025, pY= 0.026, and pV= 0.114. Set the probability of switching from the hydrophilic state to the hydrophobic state to be 0.01. Set the probability of switching from the hydrophobic state to the hydrophilic state to be 0.01. Now infer which state of the HMM is most likely to have generated each amino acid position in the the human odorant receptor 5BF1 protein (UniProt accession Q8NHC7). What is the most probable state path? The odorant receptor is a 7-transmembrane protein, meaning that it crosses the cell membrane seven times. As a consequence the protein has seven hydrophobic regions that cross the fatty cell membrane, and seven hydrophilic segments that touch the watery cytoplasm and extracellular environments. What do you think are the coordinates in the protein of the seven transmembrane regions?
