# Programming in R: for functions

**By**: Avril Coghlan  

**Adapted, edited and expanded**: Nathan Brouwer (brouwern@gmail.com) under the Creative Commons 3.0 Attribution License [(CC BY 3.0)](https://creativecommons.org/licenses/by/3.0/).

## Preface

This is a modification of ["DNA Sequence Statistics (1)"](https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html) from Avril Coghlan's [*A little book of R for bioinformatics.*](https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/index.html).  The text and code was originally written by Dr. Coghlan and distributed under the [Creative Commons 3.0](https://creativecommons.org/licenses/by/3.0/us/) license. 


## Vocab

* function
* curly brackets

## Functions

* `function()`

## Functions in R

We have been using **built-in functions** such as mean(), length(), print(), plot(), etc. We can also create our own functions in R to do calculations that you want to carry out very often on different input data sets. For example, we can create a function to calculate the value of 20 plus the square of some input number:

```{r}
myfunction <- function(x) { 
  output <- (20 + (x*x)) 
  return(output)
  }
```

This function will calculate the square of a number (x), and then add 20 to that value.  It stores this in a temporary object called output. The return() statement returns the calculated value. Once you have typed in this function, the function is then available for use. For example, we can use the function for different input numbers (e.g.. 10, 25):


```{r}
myfunction(10)
myfunction(25)
```

 
 
You can view the code that makes up a function by typing its name (without any parentheses). For example, we can try this by typing “myfunction”:

```{r}
myfunction
```
 

## Comments in R

When you are typing R, if you want to, you can write comments by writing the comment text after the “#” sign. This can be useful if you want to write some R commands that other people need to read and understand. R will ignore the comments when it is executing the commands. For example, you may want to write a comment to explain what the function log10() does:

```{r}
x <- 100
log10(x) # Finds the log to the base 10 of variable x.
```

