---
output: html_document
editor_options: 
  chunk_output_type: console
---
# How to review this book


## File format

These lessons are written in RStudio using RMarkdown.  Each .RMd file is a mix of test, written in plain format, and **code chunks**, which look like this

```{r}


```

Code chunks start with there apostrophes and {r}, like this: \`\`\`{r}.  They end with three apostrophes\`\`\`.  They will appear gray when opened up in RStudio but be white in the normal R code editor or other text editor.

## What can you do?

* Read and fix typos : )
* Add comments within the file
* Email me general comments about the files (structure, topics, confusing parts etc)

## Adding comments


You're welcome to add comments anywhere to the files, do the exercises and type up the key, propose you own exercises, etc.

### HTML-tagged comments 

The easiest way is to type a comment into the normal text part of the .Rmd file and then surround it with an html comment tag.  A comment saying "A comment" will therefore look like this:

<!-- A comment -->

In RStudio you can type up a comment, highlight it then hit Shift+Control+C on a PC or Shift+Command+C on a mac.

### Code chunk comments

Another way to add a comment is to make a RMarkdown code chunk thne type up your comments in it by commenting out each line, like the one below.  Key to doing this is to put eval = F, echo = F in the braces after the r.  eval = F, echo = F tells RStudio to leave that alone when the .Rmd file gets rendered into and web page or PDF.

```{r eval = F, echo = F}
# line 1 of comment
# line 2 of comment
# line 3 of comment.
```

In RStudio you can add code chunks with a shortcut key.  On a Mac the shortcut is OPTION + COMMAND + I.


### Keys to exercise

Some files will have exercises at the end.  I don't always include the key, and you're welcome to try the exercise and type up a key (or fix any errors in mine).  As for code chunk comments include  eval = F, echo = F in the braces so the key won't appear when rendered.    So a problem and its key would look something like this


### Problem - fix this code

```{r, eval = F}
print(correct answer)
```


```{r eval = F, echo = F}
# KEY
print("correct answer") #quotes missing
```

