# Writing functions - practice problems

<!-- TODO: find key! -->

**By**: Nathan Brouwer

## Introduction {-}

Writing functions can be a hard skill to master.  Here are some practice problems to help you build your skill and intution with them. The key is at the end.


## Pure practice

First, we'll work on some functions that don't really do anything useful.

### Hello world!

Write a function that takes no argument, and every time you run it says "Hello world!"  Call the function `hello_world()`

### Practice Function 2: Hello _______ !

Write a function that takes a word as an **argument**, and every time you run it says "Hello _____ ", where the blank is filled by the argument.  Try making the function with and without a default.  Call the function `hello()`.

**Advanced (for fun)**: Include in the function a **logical test** to see if what was passed to the argument is a string and return the message "no valid string was entered" if something other than a string was entered.


## Practice Function 3: Make you own natural log function

R uses `log()` for the natural log and `log10()` for the base-10 log.  It would be easier for non-R people to read code if there was a function with "e" in its name to indicate when the natural log is being used.  Write a function that takes a number as an argument and returns the natural log.  Call the function `log_e`

**Advanced (for fun)**: make a function called `my_log()` that has an argument "type" to specify the type of log you want.


## Practice Function 4: Make function that converts DNA directly to RNA 

The `gsub()` function can be used to replace characters in a string using **regular expressions.  For example

```{r}
str <- "ATCG"
gsub("T","U",str)


str <- "ATTCG"
gsub("T","U",str)

str <- "ATTCGTTT"
gsub("T","U",str)

str <- "AAAA"
gsub("T","U",str)
```

Create a function called `dna_to_rna()` that takes a character string as an argument and turns T to U.

## Key

### Practice Function 1: Hello world!

Write a function that takes no arguement, and every time you run it says "Hello world!"  Call the function hello_world()

```{r}
#best
hello_world <- function(){
  return("Hello world")
  }
hello_world()


# these work too
hello_world <- function(){
  print("Hello world")
  }
hello_world()



hello_world <- function(){
  cat("Hello world")
  }
hello_world()
```


### Practice Function 2: Hello _______ !

Write a function that takes a word as an argument, and every time you run it says "Hello _____ ", where the blank is filled by the argument.  Try making the function with and without a default.  Call the function hello().

```{r}
hello <- function(x){
  hello.x <- paste("Hello ",x)
  return(hello.x)}
hello("mom")

hello <- function(x){
  print("Hello")
  print(x)
  }
hello("mom")
```




## Practice Function 3: Make you own natural log function

R uses log() for natural and log10() for the base 10 log.  It would be easier for non-R peole to read code if there was a function with "e" in its name to indicate when the natural log is being used.  Write a function that takes a number as an argument and returns the natural log.  Call the function log_e

```{r}
log_e <- function(x){
  log(x)
}

log_e(10)
log(10)
```




