# Simulating random amino acid sequences

**By**: Nathan Brouwer

## Selecting random letters from a vector

Make a vector with some letters
```{r}
my.letters <- c("A","B","C","D")
```

Use sample() to select one random letter

```{r}
sample(x = my.letters, size = 1, replace = T)
```

Select 10 letters, with replacement (replace = T), so that the same letter can occur more than once

```{r}
sample(x = my.letters, size = 10, replace = T)
```

Select 100 letters
```{r}
sample(x = my.letters, size = 10, replace = T)
```


## Select random letters from the whole alphabet

The object LETTERS has the whole alphabet

```{r}
LETTERS
```


Select 10 letters from whole alphabet
```{r}
sample(x = my.letters, size = 10, replace = T)
```


## Select random amino acids to build polypeptide

Make a vector of all letters that represent and amino acid
```{r}
all.aas <- c("A","C","D","E","F","G","H","I","K","L","M","N","P","Q","R","S","T","V","W","Y")
```


Randomly select an amino acid

```{r}
sample(x = all.aas, size = 1, replace = T)
```

Make a vector of 200 amino acids
```{r, eval = F}
sample(x = all.aas, size = 200, replace = T)
```


I may want to represent the size of the fake polypeptide with an object
```{r, eval = F}
my.polypep.size <- 200
sample(x = all.aas, size = my.polypep.size, replace = T)
```

I can save this an object
```{r}
my.polypep.size <- 200
my.pp <- sample(x = all.aas, size = my.polypep.size, replace = T)
```

I can make a single character string with not spaces using paste

```{r}
paste(x = my.pp,sep = "", collapse = "")
```


```{r}

```

