---
title: "Tutorial: Writing functions to calculate the number of phylogenetic trees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: Writing functions to calculate the number of phylogenetic trees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r  eval = T}
library(compbio4all)
```




### Functions

factorial()
seq()
print()
c()
if()
function()
args()

### Concepts:

Number of possible phylogenetics trees 
vectorized inputs to functions
conditional statements
Role of arguements in writing function
Role of { and } in writing functions
function source code
functions with defaults
functions with 1 arguement
functions with 2 arguements

### Vocab:

function
function arguement
function default
conditional statement


### Packages:

None



## Introduction

Phylogenetics is one of the original fields of computational biology.  This is because as the number of taxa that need to be organized on a tree increases, the number of unique phylogenetic trees increases rapidly.

In this tutorial, you'll learn how to write functions in R so that you can calculate how many trees are possible for any number of taxa.

Prior to completing this assignment you should work through the function-writing tutorials in swirl and on RStudio Cloud.

## Number of rooted trees

The number of possible rooted phylogenetic trees is calculated using the equation below, where n is the number of taxa:

In text:
(2n-3)!/(2^(n-2))(n-2)!

Rendered:

$\frac{(2*n-3)!}{2^{n-2} * (n-2)!}$


## Number of rooted trees in R

In R we take factorials using the factorial() function
```{r}
factorial(3)
3*2*1

factorial(4)
4*3*2*1
```

We can re-write our text equation above as
factorial(2n-3)/(2^(n-2))*(factorial(n-2))

In this would be

```{r}
n<-3

factorial(2*n-3)/((2^(n-2))*factorial(n-2))
```

All of the parentheses make this a bit nutty.  Let me write this out as a separate numerator and denominator

```{r}
#numerator
numerator   <- factorial(2*n-3)

#denominator
denominator <- 2^(n-2)*factorial(n-2)

#division
numerator/denominator
```

## Functions in R

Functions in R have the general format

function_name <- function(arguements = ...){
 # math etc saved to object
 print(output)
}

The function_name can be any valid R object name.
The function function() creates the function.
A function can have any number of arguments.
Note that the arguments are enclosed in parentheses, and after the closing ")" there is a "}".
After all the stuff inside the function its finished with a "}".

## Function to calculate the number of trees

We can set up a function to encapsulate this.  We'll call the function tree_count().  It has one **argument** -- n --  the number of taxa on the tree.  We'll set the **default value** of the arguement to be n = 3 taxa.

I use the function print() at the end so that the function acutally returns output.  If this isn't included then nothing will be returned by the function.

In the code below note the locations of the following things

* function ame
* function-making command
* The opening parenthesis (
* the arguement
* the default
* the closing parenthesis )
* the opening curly bracket {
* the math
* the print() command
* the closing curly bracket }

```{r}

#             [ ]function-making command
#             |
# [ ]Function |         [ ] argument
# name        |          |
# |           |          |    [ ]closing ")"
# |           |          |     |
# |           | [ ] "("  |     |  [ ]"{"
# |           |      \   |     | /
tree_count <- function(  n = 3 ){    # note the "}"
  
  # [ ] math
  numerator   <- factorial(2*n-3)
  denominator <- 2^(n-2)*factorial(n-2)
  trees  <- numerator / denominator 
  
  # [ ] print output
  print(trees)
  
}  # end with the "}"
#\
# [ ] "}"

           
```

Does it work?
```{r}
tree_count()
```

We can view the **function source code** by just running the name of the function without any parantheses (or arugements)
```{r}
tree_count
```


We can see just the arguments, and any defaults, for a function using args()
```{r}
args(tree_count)
```


As an aside: the print() function is key.  Make and run this function and see what happens.
```{r}

                                 #/ don't forget this!
tree_count_bad <- function(n = 3){
  
  # math
  numerator   <- factorial(2*n-3)
  denominator <- 2^(n-2)*factorial(n-2)
  trees  <- numerator / denominator 
  
}
#\ don't forget this!

tree_count_bad()
```






How rapidly does the number of possible trees grow?
```{r}
tree_count(n = 3)
tree_count(n = 4)
tree_count(n = 5)
tree_count(n = 6)
```

One of my favorite example data sets has 11 taxa.
```{r}
tree_count(n = 11)
```



This number grows very rapidly.  According to Bianconi et al (2012) https://www.tandfonline.com/doi/full/10.3109/03014460.2013.807878 there are 3.72*10^13 cells in the body (37,200,000,000,000, or 3.72e13)

```{r}
### with all the zeros
37200000000000

### "e" scientific notation
3.72e13

3.72e+13
```

If we put 14 taxa on a tree there are this many possible trees
```{r}
tree_count(n = 14)
```
If we put 15 taxa on the tree
```{r}
tree_count(n = 15)
```


## Vectorized inputs to functions

One of R's most powerful features is that it works on vectors.  I want to make a plot of the the number of possible trees for 3 to 11 taxa.  I can make a vector any way I chose

```{r}
#typing out all numbers
n.taxa <- c(3,4,5,6,7,8,9,11)

# using seq with arguments spelled out
n.taxa <- seq(from = 3,to = 11,by = 1)

#using seq w/o arguments
n.taxa <- seq(3, 11, 1)

#shortcuts
n.taxa <- c(3:11)
n.taxa <- 3:11
```

I can then put the vector as an input into my function

```{r}
tree_count(n = n.taxa)
```

I can save this output to a vector
```{r}
n.trees <- tree_count(n = n.taxa)
```

And plot things.  type = "b" plots points and a line.

```{r}
plot(n.trees ~ n.taxa, type = "b")
```

## Adding conditional statements

The equation above is only valid for 2 or more taxa.  If we put in a lower number the result doesn't make sense

```{r}
tree_count(n = 0)
tree_count(n = 1)
```

We can add a **conditional statement** so that if 0 or 1 are input then an warning is given.  Conditional statements test a logical condition which, if false, can be used to throw a warning, error, etc.

I'll use the conditional statement if(n < 3) to test if the number of taxa entered into my tree_count() function is going to be valid.  The warning() function allows me to remind the user of the function of what they should  enter.

Note that the conditional statement has the logical operation in parentheses, eg (n <2), and then what to do if n < 2 is true in curly brackets

```{r}
tree_count <- function(n = 3){
  
  # conditional statement
  ## is n a valid number for using in this equaiton?
  ## if it is NOT, throw a warning
  ## if it is a valid number, skip everything in the 
  ## { } and go to the math
  if(n < 3){
    
    # warning if test is TRUE
    warning("This function is only valid for 2 or more taxa.")
  }
  
  # If test is FALSE (n = 3 or n > 3)
  ## continue with the math
  numerator   <- factorial(2*n-3)
  denominator <- 2^(n-2)*factorial(n-2)
  trees  <- numerator / denominator 
  
  # print the results
  print(trees)
}
```

## Adding additional arguments

Perhaps we want provide the option of always printing the results in scientific notation.  We can another argument to our tree_count() function called "format".  If the argument is set to "sci", "scientific" or "e", the function will change the options for how numbers are printed.

```{r}
tree_count <- function(n = 3,                # 1st arguement: number of taxa
                       format = "standard"){ # 2nd arguement: format
  
  # conditional statement
  if(n < 2){
    warning("This function is only valid for 2 or more taxa.")
  }
  
  # the math
  numerator   <- factorial(2*n-3)
  denominator <- 2^(n-2)*factorial(n-2)
  trees  <- numerator / denominator 
  
  # another conditional statement:
  ## set formatting
  if(format %in% c("sci","scientific","e")){
    options(scipen = -2,digits = 3)
  }
  
  # print the output
  print(trees)
  
  # re-set the formatting
  options(scipen = 0,digits = 7)
}
```

Check out the source code
```{r}
tree_count
```

Check out the arguements
```{r}
args(tree_count)
```


Check the output
```{r}
tree_count(n = 10)
tree_count(n = 10, format = "sci")
tree_count(n = 10, format = "scientific")
tree_count(n = 10, format = "e")
```


