---
title: "Newick format in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Newick format in R}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r  eval = T}
library(compbio4all)
```

## Introduction

**Newick format** is a common way to store information about rooted phylogenetic trees in a compact, written format.  In its standard format, Newick does not store information about branch lengths.

(Note: Newick format is briefly mentioned on page 47 of Baum & Smith Tree Thinking ).

A simple phylogenetic tree with three species can be represented as

((Human, Chimp), Gorilla)

In Newick format, **clades** are enclosed in parentheses.  So, in the example above, (Human, Chimp) means that humans and chimps are a clade and are each other's closest neighbors.  

The general mathematical rules of  parentheses apply to Newick format.  Therefore, the order within the parentheses doesn't matter; (Human, Chimp) is the same as (Chimp, Human), and ((Human, Chimp), Gorilla) is the same as (Gorilla, (Human, Chimp)).  This relates to how phylogenetic trees work and rotations can occur around nodes without changing **topology.**  This is the same as in math.  (1+2) is the same as (2+1), and ((2+1) * 2) is the same as (2*(2+1)).

If we split Gorillas into the two major groups, western Gorillas in Gabon and eastern Gorillas in the Democratic Republic of Congo, we would put them in a clade: (Gorilla-W, Gorilla-E).  We could then build up the full tree as

((Human, Chimp), (Gorilla-W, Gorilla-E))


If wanted to represent a tree with humans, chimps, gorillas and orangutan we would represent it like this:

(((Human, Chimp),Gorilla), Orangutan)

The parenthesis to the right of Gorilla indicates that Gorilla falls within the human-chimp clade. Just as in math, the number of parentheses needs to balance out.  In this case, there are three parentheses open to the right, all next to human.  There therefore needs to be 3 parentheses open to the left.

```{r}
# 321            1        2           3
# (((Human, Chimp),Gorilla), Orangutan)
```

In R, a phylogenetic tree in Newick format can be input as a character string and plotted using the ape package.

## Preliminaries 

```{r}
library(ape)
```

## 3 taxa tree in Newick

We input the tree as quoted text; note the semi colon at the end!

shorturl.at/dktQ7
```{r}
#install.packages("ape")
#install.packages("phangorn")
library(ape)
library(phangorn)
"(H,C)"
"( (H,C) , G )"
str1<- "((H, C), G );" # semi colon!
```

We then convert the tree using read.tree
```{r}
tree1 <-read.tree(text=str1)
```

We plot with plot
```{r}
plot(tree1, main = "3-taxa tree")
```




We can change the orientation of the tree by changing the order within the parentheses.  For example, if we want Gorillas to appear at the bottom of the graph

```{r}
str2<- "(G,(H, C));" # semi colon!
tree2 <-read.tree(text=str2)
plot(tree2, main = "3-taxa tree")
```


Perhaps we want humans at the top

```{r}
str3<- "(G,(C, H));" # semi colon!
tree3 <-read.tree(text=str3)
plot(tree3, main = "3-taxa tree")
```


Note that all of these trees have the same **branching order** and therefore the exact same **topology**.  (As a challenge, try to add the forth possible format of the tree that has the same topology)

TODO doesnt render
```{r , fig.height = 3, fig.width = 5, eval = F}
par(mfrow = c(2,2))
plot(tree1, main = "3-taxa tree")
plot(tree2, main = "3-taxa tree")
plot(tree3, main = "3-taxa tree")
```



## 4 taxa tree in Newick

When there are four taxa, there are two main topologies.  The first is a simple **ladder branching** typology.  This is represented 

TODO doesnt render
```{r}
par(mfrow = c(2,2), mar = c(2,2,2,2), eval = F)
str4<- "(((Human, Chimp),Gorilla), Orangutan);"
tree4 <-read.tree(text=str4)
plot(tree4, main = "4-taxa ladder tree")

```

The second major topology is when there are two major clades
```{r}
str5<- "((Human, Chimp), (Gorilla-W, Gorilla-E));"
tree5 <-read.tree(text=str5)
plot(tree5, main = "4-taxa, 2 clade tree")

```

Within each of the topologies, the order of species can be switch depending on what the evidence suggests.  For example, some researchers have proposed the humans are more closely related to Orangutans.  Their hypothesis has a **ladder branching** typology with the order of the species switched.


```{r}
str6<- "(((Human, Orangutan),Chimp), Gorilla);"
tree6 <-read.tree(text=str6)
plot(tree6, main = "")
```


## 5 taxa in Newick Format

Five taxa can occur in a number of general typologies.  

The most basic is a continuous ladder.  After Orangutans, humans next most distantly related "cousin" are the Gibbons  

TODO
```{r}
par(mfrow = c(1,1), mar = c(2,2,2,2), eval = F)
str7<- "((((Human, Chimp),Gorilla), Orangutan),Gibbons);"
tree7 <-read.tree(text=str7)
plot(tree7, main = "5-taxa ladder tree")
```


Let's say there's a hypothesis that Gibbons and Orangutans form a clade. This typology would create a tree with two branches, one with 3 species and one with two.


```{r}

str8 <- "((((Human, Chimp),Gorilla), (Orangutan,Gibbons)));"
tree8 <-read.tree(text=str8)
plot(tree8, main = "5-taxa, 2 big clades tree")
```


A third major typology occurs if there are four taxa in a clade on a branch.  Let's say there's a hypothesis that Orangutans and Gorillas are sister species, and Gibbons are there more distantly related cousins.
```{r}

str9 <- "(((Human, Chimp), (Orangutan, Gorilla)), Gibbons);"
tree9 <-read.tree(text=str8)
plot(tree9, main = "5-taxa, 1 big clade ")
```



Here's another example of this format, dropping Gibbons and adding the two Gorilla subspecies

```{r}
str9<- "(((Human, Chimp),(Gorilla-W, Gorilla-E)), Orangutan);"
tree9 <-read.tree(text=str9)
plot(tree9, main = "")
```




## Many-taxa tree

Newick format can be used to represent any kind of tree, though in practice this can be very difficult to write out by hand.  Here is an example of a number of vertebrates

```{r}
par(mfrow = c(1,1), mar = c(2,2,2,2))
str.vert<-    "(((((((cow, pig),whale),(bat,(lemur,human))),(robin,iguana)),coelacanth),gold_fish),shark);"
vert.tree<-read.tree(text=str.vert)
plot(vert.tree,no.margin=TRUE,edge.width=2)
```


## Newick and Branch lengths

Below is an example of a BIG tree which also has branch lengths represented.  The original analysis was in  Miyamoto and Goodman (1986) "Biomolecular Systematics of Eutherian Mammals: Phylogenetic Patterns and Classification."
```{r}
str.vert<-  "(Edentata:55, (((Orycteropus:12, Trichechus:43):1, (Procavia:29, (Elephas:18, Loxodonta:5):55):10):15, (((Chiroptera:27, (Tupaia:43, ((Talpa:24, (Suncus:24, Erinaceus:58):6):4, (Manis:5, ((Felis:13, Leo:7):32, ((Canis:37, Ursidae:12):4, ((Phocidae:19, Zalophus:17):7, (Procyonidae:12, Mustelidae:22):9):5):17):13):3):10):6):8, (((Lemuridae:46, (Galago:16, Nycticebus:27):10):8, (Tarsius:15, ((Cebus:10, (Atelinae:5, (Aotus:6, Callithrix:9):3):3):18, ((Hylobates:6, (Pongo:1, (Gorilla:2, (Homo:22, Pan:1):2):6):2):5, (Presbytis:1, (Cercopithecus:1, (Erythrocebus:0, ('Macaca fascicularis':2, ('Macaca mulatta':5, 'Macaca fuscata':0):1, (Theropithecus:2, Papio:4):15):2):1):8):6):9):22):10):13, ((Ochotona:7, Oryctolagus:54):4, (Caviomorpha:107, (Spermophilus:29, (Spalax:23, ((Rattus:71, Mus:19):15, (Ondatra:27, Mesocricetus:32):27):20):8):12):15):11):9):12, ((Sus:50, ((Lama:10, Camelus:24):31, (Hippopotamus:31, (((Ovis:9, Capra:8):19, ((Antilocapra:13, Giraffa:14):7, (Cervus:8, Alces:9):11):2):9, (Tragelaphinae:6, ('Bos grunniens':6, (Bison:5, 'Bos taurus':15):7):11):6):47):10):19):18, ((('Equus (Asinus)':6, 'Equus caballus':31):24, (Tapirus:19, Rhinocerotidae:23):11):22, ((Phocoena:4, Tursiops:17):5, (Balaenoptera:36, Eschrichtius:2):8):29):12):12):16):55);"
vert.tree<-read.tree(text=str.vert)
plot(vert.tree,no.margin=TRUE,edge.width=2)
```



## Additional Reading

The Wikipedia article on Newick format shows how it can be extended for more complex scenarios, including annotation of branch lengths.
https://en.wikipedia.org/wiki/Newick_format
