# Function practice

## Part 1: Pure practice

### Practice Function 1: Hello world!

Write a function that takes no arguement, and every time you run it says "Hello world!"  Call the function hello_world()

```{r}
#best
hello_world <- function(){
  return("Hello world")
  }
hello_world()


# these work too
hello_world <- function(){
  print("Hello world")
  }
hello_world()



hello_world <- function(){
  cat("Hello world")
  }
hello_world()
```


### Practice Function 2: Hello _______ !

Write a function that takes a word as an argument, and every time you run it says "Hello _____ ", where the blank is filled by the argument.  Try making the function with and without a default.  Call the function hello().

```{r}
hello <- function(x){
  hello.x <- paste("Hello ",x)
  return(hello.x)}
hello("mom")

hello <- function(x){
  print("Hello")
  print(x)
  }
hello("mom")
```




## Practice Function 3: Make you own natural log function

R uses log() for natural and log10() for the base 10 log.  It would be easier for non-R peole to read code if there was a function with "e" in its name to indicate when the natural log is being used.  Write a function that takes a number as an argument and returns the natural log.  Call the function log_e

```{r}
log_e <- function(x){
  log(x)
}

log_e(10)
log(10)
```

