# Writing functions - practice problems

```{r  eval = T}
library(compbio4all)
```



## Introduction

Writing functions can be a hard skill to master.  Here are some practice problems.  A key is available.

## Part 1: Pure practice

### Practice Function 1: Hello world!

Write a function that takes no arguement, and every time you run it says "Hello world!"  Call the function hello_world()

### Practice Function 2: Hello _______ !

Write a function that takes a word as an argument, and every time you run it says "Hello _____ ", where the blank is filled by the argument.  Try making the function with and without a default.  Call the function hello().

**Advanced (for fun)**: Include in the function a test to see if what was passed to the arguement is a string and return the message "no valid string was entered" if something other than a string was entered.


## Practice Function 3: Make you own natural log function

R uses log() for natural and log10() for the base 10 log.  It would be easier for non-R peole to read code if there was a function with "e" in its name to indicate when the natural log is being used.  Write a function that takes a number as an argument and returns the natural log.  Call the function log_e

**Advanced (for fun)**: make a function called my_log() that has an arguement "type" to specify the type of log you want.


## Practice Function 4: Make function that converts DNA directly to RNA 

The gsub() function can be used to replace characters in a string.  For example

```{r}
str <- "ATCG"
gsub("T","U",str)


str <- "ATTCG"
gsub("T","U",str)

str <- "ATTCGTTT"
gsub("T","U",str)

str <- "AAAA"
gsub("T","U",str)
```

Create a function called dna_to_rna that takes a character string as an arguement and turns T to U.



