# Computational Biology Portfolio: Practice Working with Strings

```{r  eval = T}
library(compbio4all)
```






## Introduction

Gonder et al. (2007) published phylogenetic analyses of chimp population using the **XX Name of gene / sequence  XX** sequence from the chimp **XX Type of DNA - nuclear? mitochondrial? both XX**.  Gonder et al (2007) used both previously published and newly sequenced data to build their phylogenetic trees.  Chimp sequences were compared to **XX what was the outgroup?**.

## Purpose

**XX Write a brief summary of what this script does XX**.


## Gonder et al 2007 data

Gonder et al report their **XX What are these serial number thingy's called? XX** in two paragraphs.

First, the previously published data they used was described as:

> "We deposited all previously unpublished sequences in GenBank under the accession numbers DQ140188-DQ140269." (Gonder et al 2006 pg 1110)

Second, the new sequences they worked up were described as:

> "We obtained chimpanzee HVRI sequences from other studies from Gen-
Bank under the accession numbers L35381–L35443, U77186–U77293,
AFO59042–AFO59052, and AF137481–AF137412. We obtained human
HVRI sequences from the Human Mitochondrial Genome Database
(mtDB) at http://www.genpat.uu.se/mtDB/sequences.html. We aligned
sequences from GenBank and mtDB with reference to sequences produced
for the study. Sequences we produced and used correspond to bases
16,048–16,391 of the human mtDNA Cambridge Reference Sequence
(Andrews et al., 1999)."  (Gonder et al 2006 pg 1111)


## Gonder et al 2007 new sequences


The **XX What are these serial number thingy's called? XX** for their new sequences were reported as having the range:

DQ140188-DQ140269

This includes **XX How many sequences XX?** sequences.  **XX Describe what this code does in 1 to 2 sentences XX**

### Creating the codes

First, let me prepare some things

```{r}
## PREP

## The first part of the code: letters
prefix       <- "DQ"

## The second part of the code:
### Numeric
### start
suffix.start <- 140188

### end
suffix.end   <- 140269
```

Now I'll re-create the actual codes using seq() and paste()
```{r}
## RECREATE CODES

## seq()
## Use the seq() command to generate the whole range
### of numeric values
suffix <- seq(from = suffix.start, to = suffix.end, by = 1)

## How many codes?
length(suffix)

## paste()
## Use paste with sep = ""
### 
DQ.codes <- paste(prefix, suffix, sep = "")

## check length
length(DQ.codes)
```

### Selecting a code for demonstration

The grep() function is a **XX What is the general name for the type of function grep() is? XX**?  I can use it to select particular elements from a vector containing text or numbers.  grep() is a difficult command to master and its VERY picky about its input.  I'm just going to demonstrate what it can do here - you'll need a lot of practice to get good with it.

```{r}
# Day of the month of was born
## as CHARACTER data
day.of.month.born <- "21"


# What grep() does
grep(pattern = day.of.month.born, # pattern to search for
     x = DQ.codes)                   # where to search

# storing grep() output
my.DQ.index <- grep(pattern = day.of.month.born, 
     x = DQ.codes)  
```


Locating a code to use
```{r}
DQ.codes[my.DQ.index]
```

I am going to use this code
"DQ140221"

When I enter it into the NCBI website I will drop the quotation marks

DQ140221


## Searching Genbank

I searched the NCBI databases for this code (DQ140221) from this website

https://www.ncbi.nlm.nih.gov/

NCBI returned this summary

https://www.ncbi.nlm.nih.gov/search/all/?term=DQ140221

From the NCBI search results I search Genbank, which returned this entry.

https://www.ncbi.nlm.nih.gov/nuccore/DQ140221.1/

* The sequence is from **Pan troglodytes** with no subspecies specified.
* The sequence is **344 bases long**
* The authors are **Gonder et al. (unpublished)**
* NCBI indicates that there are no publications linked to this entry.
* NCBI indicates there is **1 "popset"** linked to this entry.  **The PopSet is available at https://www.ncbi.nlm.nih.gov/popset/?term=DQ140221**



## Gonder et al 2007 previously published sequences

The previously published sequences had the following code:

* L35381–L35443 
* U77186–U77293
* AF059042–AF059052
* AF137481–AF137412

I selected the **XX what series? XX** series because **XX What did you select it?XX**.  This series has **XX after you work things up use length() to determine this number XX** codes in it.

I then generate all the codes in the series and selected one that contained the day I was born.


### Creating the codes

First, let me prepare some things

```{r}
## PREP

## The first part of the code: letters
                      ##
prefix       <- "DQ"  ##1) Change to your pick, L, U, or AF

## The second part of the code:
### Numeric
### start
                       ##
suffix.start <- 140188 ##2a) Change to your pick
                       ## MAKE SURE IT MATCHES THE PREFIX!

### end
                       ##
                       ##2b) Change to your pick
                       ## MAKE SURE IT MATCHES THE PREFIX
suffix.end   <- 140269
```

Now I'll re-create the actual codes using seq() and paste()
```{r}
## RECREATE CODES

## seq()
## Use the seq() command to generate the whole range
### of numeric values
suffix <- seq(from = suffix.start, 
              to = suffix.end, 
              by = 1)

## How many codes?
              ##3)determine this and add it to the notes above


## paste()
## Use paste with sep = ""
### 
DQ.codes <- paste(prefix, 
                  suffix, 
                  sep = "")

## check length


```

### Selecting a code for demonstration


```{r}
# Day of the month of was born
## as CHARACTER data
                       ##4) Change to a number 1 to 31
day.of.month.born <- "21"


# What grep() does
grep(pattern = day.of.month.born, # pattern to search for
     x = DQ.codes)                # where to search

# storing grep() output
my.DQ.index <- grep(pattern = day.of.month.born, 
     x = DQ.codes)  
```


Locating a code to use
```{r}
DQ.codes[my.DQ.index]
```

I am going to use this code
""   

When I enter it into the NCBI website I will drop the quotation marks




## Searching Genbank

I searched the NCBI databases for this code (**## YOUR CODE HERE##**) from this website

https://www.ncbi.nlm.nih.gov/

NCBI returned this summary

**## YOUR summary HERE##**

From the NCBI search results I search Genbank, which returned this entry.

**## YOUR genebank entry HERE##**

* The sequence is from **##YOUR SPECIES w/SUBSPECIES if listed##** with no subspecies specified.
* The sequence is **##YOUR SEQUENCE LENGTH##**
* The authors are **##YOUR AUTHORS HERE##**
* NCBI indicates **##NUMBER OF PUBLICATIONS IF ANY LINKED##**.
* NCBI indicates there is **##NUMBER OF POPSETS IF ANY ##** linked to this entry.  **##LINK TO POPSET##**
