---
title: "ac-mv) Standardizing Variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ac-mv) Standardizing Variables}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = F}
library(compbio4all)
```

Emily Jennison

## Standardizing Variables

If you want to compare different variables that have different units, are very different variances, it is a good idea to first standardize the variables.

For example, we found above that the concentrations of the 13 chemicals in the wine samples show a wide range of standard deviations, from 0.1244533 for V9 (variance 0.01548862) to 314.9074743 for V14 (variance 99166.72). This is a range of approximately 6,402,554-fold in the variances.

As a result, it is not a good idea to use the unstandardized chemical concentrations as the input for a principal component analysis (PCA, see below) of the wine samples, as if you did that, the first principal component would be dominated by the variables which show the largest variances, such as V14.

Thus, it would be a better idea to first standardize the variables so that they all have variance 1 and mean 0, and to then carry out the principal component analysis on the standardized data. This would allow us to find the principal components that provide the best low-dimensional representation of the variation in the original data, without being overly biased by those variables that show the most variance in the original data.

You can standardize variables in R using the “scale()” function.

```{r}
data(wine)
```


For example, to standardize the concentrations of the 13 chemicals in the wine samples, we type:

```{r}
standardisedconcentrations <- as.data.frame(scale(wine[2:14]))
```

Note that we use the “as.data.frame()” function to convert the output of “scale()” into a “data frame”, which is the same type of R variable that the “wine” variable.

We can check that each of the standardized variables stored in “standardisedconcentrations” has a mean of 0 and a standard deviation of 1 by typing:

```{r}
sapply(standardisedconcentrations,mean)

sapply(standardisedconcentrations,sd)

```

We see that the means of the standardized variables are all very tiny numbers and so are essentially equal to 0, and the standard deviations of the standardized variables are all equal to 1.
