---
output: html_document
editor_options: 
  chunk_output_type: console
---
# A primer for working with vecotrs

**By**: Avril Coghlan  

**Adapted, edited and expanded**: Nathan Brouwer (brouwern@gmail.com) under the Creative Commons 3.0 Attribution License [(CC BY 3.0)](https://creativecommons.org/licenses/by/3.0/).


## Preface 

<!-- rename this attribution?  put at end? -->

This is a modification of part of["DNA Sequence Statistics (2)"](https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter2.html) from Avril Coghlan's [*A little book of R for bioinformatics.*](https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/index.html).  Most of text and code was originally written by Dr. Coghlan and distributed under the [Creative Commons 3.0](https://creativecommons.org/licenses/by/3.0/us/) license. 

## Vocab 

* base R
* scalar, vector, matrix
* vectorized operation
* regular expressions

# Functions
* `seq()`
* `is()`, `is.vector()`, `is.matrix()`
* `gsub()`




## Vectors in R

**Variables** in R include **scalars**, **vectors**, and **lists**.  **Functions** in R carry out operations on variables, for example, using the `log10()` function to calculate the log to the base 10 of a scalar variable `x`, or using the `mean()` function to calculate the average of the values in a vector variable `myvector`.  For example, we can use `log10()` on a scalar object like this:

```{r}
# store value in object
x <- 100

# take log base 10 of object
log10(x)
```

Note that while mathematically x is a single number, or a scalar, R considers it to be a vector:

```{r}
is.vector(x)
```

There are many "is" commands.  What is returned when you run `is.matrix()` on a vector?
```{r}
is.matrix(x)
```


Mathematically this is a bit odd, since often a vector is defined as a one-dimensional matrix, e.g., a single column or single row  of a matrix.  But in *R* land, a vector is a vector, and matrix is a matrix, and there are no explicit scalars.


## Math on vectors

Vectors can serve as the input for mathematical operations. When this is done *R* does the mathematical operation separately on each element of the vector.  This is a unique feature of *R* that can be hard to get used to even for people with previous programming experience.

Let's make a vector of numbers:

```{r}
myvector <- c(30,16,303,99,11,111)
```

What happens when we multiply `myvector` by 10?

```{r}
myvector*10
```

R has taken each of the 6 values, 30 through 111, of `myvector` and multiplied each one by 10, giving us 6 results.  That is, what R did was

```{r, eval =-F}
30*10    # first value of myvector
16*10    # second value of myvector
303*10   # ....
99*10
111*10   # last value of myvector
```

 The normal order of operations rules apply to vectors as they do to operations we're more used to.   So multiplying `myvector` by 10 is the same whether you put he 10 before or after vector.  That is `myvector\*10` is the same as `10\*myvector`.

```{r}
myvector*10
10*myvector
```


What happen when you subtract 30 from myvector?  Write the code below.

```{r}
myvector-30
```

So, what R did was
```{r, eval =-F}
30-30    # first value of myvector
16-30    # second value of myvector
303-30   # ....
99-30
111-30   # last value of myvector
```

Again, `myvector-30` is vectorized operation.

You can also square a vector

```{r}
myvector^2
```

Which is the same as
```{r, eval =-F}
30^2    # first value of myvector
16^2    # second value of myvector
303^2   # ....
99^2
111^2   # last value of myvector
```

Also you can take the square root of a vector using the functions `sqrt()`...

```{r}
sqrt(myvector)
```


...and take the log of a vector with `log()`...

```{r}

log(myvector)

```

...and just about any other mathematical operation.  Here we are working on a separate vector object; all of these rules apply to a column in a matrix or a dataframe.  

This attribute of R is called **vectorization**.  When you run the code `myvector*10` or `log(myvector)` you are doing a **vectorized** operation - its like normal math with special vector-based super power to get more done faster than you normally could.


## Functions on vectors

As we just saw, we can use functions on vectors.  Typically these use the vectors as an input and all the numbers are processed into an output.  Call the `mean()` function  on the vector we made called `myvector`.
```{r}
mean(myvector)
```
 
Note how we get a single value back - the mean of all the values in the vector.  R saw that we had a vector of multiple and knew that the mean is a function that doesn't get applied to single number, but sets of numbers.
 
The function `sd()` calculates the standard deviation. Apply the `sd()` to myvector:

```{r}
sd(myvector)
```


## Operations with two vectors

You can also subtract one vector from another vector.  This can be a little weird when you first see it.  Make another vector with the numbers 5, 10, 15, 20, 25, 30.  Call this myvector2:

```{r}
myvector2 <- c(5, 10, 15, 20, 25, 30)
```


Now subtract myvector2 from myvector.  What happens?

```{r}
myvector-myvector2
```

 
## Subsetting vectors
 
You can extract an **element** of a vector by typing the vector name with the index of that element given in **square brackets**. For example, to get the value of the 3rd element in the vector `myvector`, we type:


```{r}
myvector[3]
```


Extract the 4th element of the vector:

```{r}
myvector[4]
```


You can extract more than one element by using a vector in the brackets:

First, say I want to extract the 3rd and the 4th element.  I can make a vector with 3 and 4 in it:
```{r}
nums <- c(3,4)
```


Then put that vector in the brackets:
```{r}
myvector[nums]
```


We can also do it directly like this, skipping the vector-creation step:

```{r}
myvector[c(3,4)]
```


In the chunk below extract the 1st and 2nd elements:

```{r}
myvector[c(1,2)]
```

## Sequences of numbers

Often we want a vector of numbers in **sequential order**. That is, a vector with the numbers 1, 2, 3, 4, ... or 5, 10, 15, 20, ...  The easiest way to do this is using a colon
```{r}
1:10
```

Note that in R 1:10 is equivalent to c(1:10)
```{r}
c(1:10)
```

Usually to emphasize that a vector is being created I will use c(1:10)

We can do any number to any numbers
```{r}
c(20:30)
```


We can also do it in *reverse*.  In the code below put 30 before 20:

```{r}
c(30:20)
```


A useful function in *R* is the `seq()` function, which is an explicit function that can be used to create a vector containing a sequence of numbers that run from a particular number to another particular number.

```{r}
seq(1, 10)
```

Using `seq()` instead of a `:` can be useful for readability to make it explicit what is going on.  More importantly, `seq` has an argument `by = ...` so you can make a sequence of number with any interval between  For example, if we want to create the sequence of numbers from 1 to 10 in steps of 1 (i.e.. 1, 2, 3, 4, ... 10), we can type:

```{r}
seq(1, 10,
    by = 1)
```


We can change the **step size** by altering the value of the `by` argument given to the function `seq()`. For example, if we want to create a sequence of numbers from 1-100 in steps of 20 (i.e.. 1, 21, 41, ... 101), we can type:

```{r}
seq(1, 101,
    by = 20)
```



## Vectors can hold numeric or character data

The vector we created above holds numeric data, as indicated by `class()`

```{r}
class(myvector)
```

Vectors can also holder character data, like the genetic code:

```{r}
# vector of character data
myvector <- c("A","T","G")

# how it looks
myvector

# what is "is"
class(myvector)
```

## Regular expressions can modify character data

We can use **regular expressions** to modify character data. For example, change the Ts to Us

```{r}
myvector <- gsub("T", "U", myvector)
```

Now check it out
```{r}
myvector
```

Regular expressions are a deep subject in computing.  You can find some more information about them [here](https://rstudio-pubs-static.s3.amazonaws.com/74603_76cd14d5983f47408fdf0b323550b846.html).


