---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Plotting vectors in base R {#plotting-vectors}

**By**: Avril Coghlan  

**Adapted, edited and expanded**: Nathan Brouwer (brouwern@gmail.com) under the Creative Commons 3.0 Attribution License [(CC BY 3.0)](https://creativecommons.org/licenses/by/3.0/).


## Preface 

<!-- rename this attribution?  put at end? -->

This is a modification of part of["DNA Sequence Statistics (2)"](https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter2.html) from Avril Coghlan's [*A little book of R for bioinformatics.*](https://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/index.html).  Most of text and code was originally written by Dr. Coghlan and distributed under the [Creative Commons 3.0](https://creativecommons.org/licenses/by/3.0/us/) license. 


## Plotting numeric data

R allows the production of a variety of plots, including **scatterplots**, **histograms**, **piecharts**, and **boxplots**.  Usually we make plots from **dataframes** with 2 or more columns, but we can also make them from two separate vectors.  This flexibility is useful, but also can cause some confusion.   

For example, if you have two equal-length vectors of numbers `numeric.vect1` and `numeric.vect1`, you can plot a [**scatterplot**](https://en.wikipedia.org/wiki/Scatter_plot) of the values in `myvector1` against the values in `myvector2` using the **base R** `plot() `function. 

First, let's make up some data in put it in vectors:

```{r}
numeric.vect1 <- c(10, 15,  22,  35,  43)
numeric.vect2 <- c(3,  3.2, 3.9, 4.1, 5.2)
```

Not plot with the base R `plot()` function:
```{r}
plot(numeric.vect1, numeric.vect2)
```

Note that there is a comma between the two vector names.  When building plots from dataframes you usually see a tilde (~), but when you have two vectors you can use just a comma.  

Also note the order of the vectors within the `plot()` command and which axes they appear on.  The first vector is `numeric.vect1` and it appears on the horizontal x-axis.

If you want to label the axes on the plot, you can do this by giving the `plot()` function values for its optional arguments `xlab = ` and `ylab = `:
```{r}
plot(numeric.vect1,   # note again the comma, not a ~
     numeric.vect2, 
     xlab="vector1", 
     ylab="vector2")
```

We can store character data in vectors so if we want we could do this to set up our labels:
```{r}
mylabels <-  c("numeric.vect1","numeric.vect2")
```

Then use bracket notation to call the labels from the vector
```{r}
plot(numeric.vect1, 
     numeric.vect2, 
     xlab=mylabels[1],
     ylab=mylabels[2])

```


If we want we can use a tilde to make our plot like this:
```{r}
plot(numeric.vect2 ~ numeric.vect1)
```

Note that now, `numeric.vect2` is on the left and `numeric.vect1` is on the right.  This flexibility can be tricky to keep track of.

We can also combine these vectors into a dataframe and plot the data by referencing the data frame.  First, we combine the two separate vectors into a dataframe using the `cbind()` command.

```{r}
df <- cbind(numeric.vect1, numeric.vect2)
```

Then we plot it like this, referencing the dataframe `df` via the `data = ...` argument.

```{r}
plot(numeric.vect2 ~ numeric.vect1, data = df)
```


## Other plotting packages

Base R has lots of plotting functions; additionally, people have written packages to implement new plotting capabilities. The package `ggplot2` is currently the most popular plotting package, and `ggpubr` is a package which makes `ggplot2` easier to use.  For quick plots we'll use base R functions, and when we get to more important things we'll use ggplot2 and ggpubr.

