# Intro to R objects {#R-objects}

**By**: Nathan Brouwer

## Commands used

* <-
* c()
* length()
* dim()
* is()


## R Objects

* Everything in R is an object, works with an object, tells you about an object, etc
* We'll do a simple data analysis with a t.test and then look at properties of R objects
* There are several types of objects: **vectors, matrices, lists, dataframes**
* R objects can hold numbers, text, or both
* A typical dataframe has columns of **numeric data** and columns of text that represent **factor variables** (aka "**categorical variables**")



## Differences between objects

Different objects are used and show up in different contexts.

* Most practical stats work in R is done with **dataframes** .  
* A dataframe is kind of like a spreadsheet, loaded into R.
* For the sake of simplicity, we often load data in as a **vector**.  This just makes things smoother when we are starting out.
* **vectors** pop up in many places, usually in a support role until you start doing more programming.
* **matrices** are occasionally used for applied stats stuff but show up more for programming.  A matrix is like a stripped-down dataframe.
* **lists** show up everywhere, but you often don't know it; many R functions make lists
* Understanding **lists** will help you efficiently work with stats output and make plots.




## The Data

We'll use the following data to explore R objects.

Motulsky 2nd Ed, Chapter 30, page 220, Table 30.1.  Maximal relaxation of muscle strips of old and young rat bladders stimulated with high concentrations of [norepinephrine](https://en.wikipedia.org/wiki/Norepinephrine) (Frazier et al. 2006).  Response variable is % E.max

```{r, echo = F}

#Old data
old.E.max <- c(20.8,2.8,50.0,33.3,29.4,38.9, 29.4,52.6,14.3)

#Young data
young.E.max <- c(45.5,55.0, 60.7, 61.5, 61.1, 65.5,42.9,37.5)

```



## The assignment operator "<-" makes object

The code above has made two objects.  We can use several commands to learn about these objects.

* is(): what an object is, i.e., vector, matrix, list, dataframe
* length():how long an object is; only works with vectors and lists, not dataframes!
* dim(): how long AND how wide and object is; doesn't work with vectors, only dataframes and matrices :(


### is()

What is our "old.E.max" object?

```{r}

is(old.E.max)
is(young.E.max)
```



Its a vector, containing numeric data.


What if we made a vector like this?
```{r}
cat.variables <- c("old","old","old","old",
                   "old","old","old","old","old")
```



And used `is()`

```{r}
is(cat.variables)
```

It tells us we have a vector, containing character data.  Not sure why it feels the need to tell us all the other stuff...


### length()

Our vector has 9 elements, or is 9 elements long.
```{r}
length(old.E.max)
```



Note that `dim()`, for dimension, doesn't work with vectors!

```{r}
dim(old.E.max)
```

It would be nice if it said something like "1 x 9" for 1 row tall and 9 elements long.  So it goes.


### str()

`str()` stands for "structure". 

* It summarizes info about an object; 
* I find it most useful for looking at lists.  
* If our vector here was really really long, str() would only show the first part of the vector

```{r}
str(old.E.max)
```



### c()

* We typically use `c()` to gather together things like numbers, as we did to make our objects above. 
* note: this is *lower case* "c"! 
* Uppercase is something else
* For me, R's font makes it hard sometimes to tell the difference between "c" and "C"
* If code isn't working, one problem might be a "C" instead of a "c"

Use `c()` to combine two objects

```{r}
old.plus.new <- c(old.E.max, young.E.max)
```



Look at the length

```{r}
length(old.plus.new)
```



Note that `str()` just shows us the first few digits, not all 17
```{r}
str(old.plus.new)
```




## Debrief

We can...

* learn about objects using length(), is(), str()
* access parts of list using $ (and also brackets)
* access parts of vectors using square brackets [ ]
* save the output of a model / test to an object
* access part of lists for plotting instead of copying stuff
